---
title: "An Algorithm to Automate the Adjudication of Lines of Treatment (LOT) for Multiple Myeloma"
author: "Valerie Pan"
date: "2026-04-06"
output:
  html_document:
    toc: true
  pdf_document: 
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Data Cleaning 

```{r load_packages, warning=FALSE, message=FALSE}
# code to install packages if needed
#if (!require(readxl)) install.packages("readxl") 
#if (!require(tidyverse)) install.packages("tidyverse") # install package if needed

library(readxl)
library(tidyverse)
```

This is code specific for cleaning **FlatIron** Data. The other dataset may need to be cleaned before applying the LOT Algorithm.  

```{r load_data, message = FALSE}
df <- read_excel("LoT Adjudication Datasets.xlsx", col_names = FALSE, sheet = 1, skip = 3)
# don't use first line as header, and skip 3 first lines (which are header in excel sheet)
head(df) # look at the data
```

```{r clean_data}
# rename the columns (unsure what column 17 and 18 refer to)
names(df)[1:(length(df)-2)] <- c("f", "Diagnosis Date", "Death Flag", "Death Date", 
                             "Treatment Regimens", "Reviewer 1 Name", 
                             "Reviewer 1 LoT", "Reviewer 2 Name", 
                             "Reviewer 2 LoT", "Flatiron Line Number", 
                             "startdate", "enddate", "Response", 
                             "Assessment Date", "Specimen Type", "residual")

df_lot <- df %>% select(c("f", "Diagnosis Date", "Death Date", 
                          "Treatment Regimens", "Reviewer 1 LoT", 
                          "Reviewer 2 LoT", "Flatiron Line Number"))

head(df_lot) # look at the data
```

```{r tester_NA}
# observe what the NA rows look like 
na <- data.frame()
for (i in 1:length(df_lot$`f`)) {
  row <- df_lot[i, ]
  
  if (is.na(row$`f`)) { 
    na <- rbind(na, row)
  }
}
```

From the test loop above, we see that in all cases where 'f' is NA, all other columns except 'trt' are NA. We can therefore assume that it will be a continuation of the previous row's 'trt', unless 'trt' is also NA. 

In the following code block, we will mutate the dataframe df_lot into a dataframe df_patient. 
df_patient is further organized, with each 'f' (patient id) appearing only once in the dataframe as a row. All treatment regimens for a given 'f' are saved in a vector that is found in the 'trt' column. 
Reviewer 1 and 2 LOT will be saved in the columns 'rev1_lot' and 'rev2_lot' to be used in later evaluation of our LOT algorithm. We also save an additional column, count_row, which is the original dataset's LOT count. This will also be compared to our LOT algorithm later.  

```{r mutate_data_new, warning = FALSE}

# initialization 
f <- df_lot[1, ]$`f`  # initialize f (first patient)
trt <- c()  # intialize a vector 
rev1_lot <- as.integer(df_lot[1, ]$`Reviewer 1 LoT`) # initialize reviewer 1's LOT
rev2_lot <- as.integer(df_lot[1, ]$`Reviewer 2 LoT`) # initialize reviewer 2's LOT
count_row <- as.integer(df_lot[1, ]$`Flatiron Line Number`) # initialize data's LOT count

# initialize an empty data frame
df_patient <- tibble(f = character(), trt = list(), rev1_lot = integer(), 
                     rev2_lot = integer(), count_row = integer()) 

# go through each row in df_lot
# organize the data further in a new dataset df_patient
for(i in 1:length(df_lot$`f`)) { 
  
  row <- df_lot[i, ] # the current row 
  
  if ( is.na(row$`f`) ) { 
    # 'f' is NA and 'trt' is not NA
    # don't change 'f' but the 'trt' belongs to the previous line 
    # (by our logic from the NA test loop)
    
    # save the 'trt' string of the 'f' = NA row to the previous row's 'trt'
    if (is.na(row$`Treatment Regimens`) == FALSE){
      trt[length(trt)] <- paste0(trt[length(trt)], df_lot[i, ]$`Treatment Regimens`)
    }
    
  } else if (row$`f` == f) { # current row f matches f 
    
    f <- df_lot[i, ]$`f` # the current patient
    
    # save the reviewer's LOTs
    rev1_lot <- as.integer(df_lot[i, ]$`Reviewer 1 LoT`)
    rev2_lot <- as.integer(df_lot[i, ]$`Reviewer 2 LoT`)
    count_row <- as.integer(df_lot[i, ]$`Flatiron Line Number`)
    
    # store all the treatments for a single patient in a vector
    trt <- c(trt, df_lot[i, ]$`Treatment Regimens`)
    
    # only when the last row is reached
    if (i == length(df_lot$`f`)) { 
      # saving the last data into the dataframe
      
      # save 'f' and 'trt' to a df 
      df_patient <- df_patient %>% add_row(f = f, trt = list(trt), rev1_lot = rev1_lot, 
                                           rev2_lot = rev2_lot, count_row = count_row)  
    
      f <- df_lot[i, ]$`f` # save the new patient as current patient
      rev1_lot <- as.integer(df_lot[i, ]$`Reviewer 1 LoT`)
      rev2_lot <- as.integer(df_lot[i, ]$`Reviewer 2 LoT`)
      count_row <- as.integer(df_lot[i, ]$`Flatiron Line Number`)
      trt <- c() # reset trt, and then add the new one
      trt <- c(trt, df_lot[i, ]$`Treatment Regimens`)
      
    } 
    
  } else if (row$`f` != f) { # row changes (to a new 'f'/ id)
    # save previous f and trt to a df 
    df_patient <- df_patient %>% add_row(f = f, trt = list(trt), rev1_lot = rev1_lot, 
                                         rev2_lot = rev2_lot, count_row = count_row)
    
    f <- df_lot[i, ]$`f` # save the new patient as current patient
    rev1_lot <- as.integer(df_lot[i, ]$`Reviewer 1 LoT`)
    rev2_lot <- as.integer(df_lot[i, ]$`Reviewer 2 LoT`)
    count_row <- as.integer(df_lot[i, ]$`Flatiron Line Number`)
    trt <- c() # reset trt, and then add the new one
    trt <- c(trt, df_lot[i, ]$`Treatment Regimens`)
    
  }
}
```

```{r check}
# examples to show we correctly identified and saved the trts for each patient (f)
df_patient[1, ]; df_patient[1, ]$trt

# further examples: 
# df_patient[2, ]; df_patient[2, ]$trt
# df_patient[110, ]; df_patient[110, ]$trt 
```

Data cleaning is successful, so I will now proceed with codifying the rules. 

\newpage

# Defining a Line of Therapy (LOT)

**A New Line of Therapy:** any one of the following conditions met

1. start a new line of treatment after discontinuing a previous line 

    - regimen is not discontinued if some drugs of regimen are not discontinued

2. **unplanned** addition or substitution (switching) of 1+ drugs in a regimen is a new line of therapy 

3. in patients undergoing >1 SCT, each SCT (autologous or allogenic) is a new line of therapy

    - Although this rule 3 is here, it seems like "Transplant" doesn't count as a new line of therapy. 

```{r tester_Transplant}
transplant <- data.frame()

# observe what the transplant rows look like
for (i in 1:nrow(df_patient)) {
  patient <- df_patient[i, ]
  trt <- patient$trt[[1]] # extract vector of trt in a row
  
  if ("Transplant" %in% trt) { 
     transplant <- rbind(transplant, patient)
  }
  
}
```

From the test code above, we note that when "Transplant" is observed in the dataset, it is not counted as a new LOT (despite the rule for SCT). There is a rule in the algorithm below to account for this.

# The LOT Algorithm

```{r lot_function}
# function automating the adjudication of lot 
lot_algorithm <- function(patient){ # patient is a row of the dataframe
  
  # initialization
  trt <- patient$trt[[1]] # extract vector of trts
  
  if (any("Transplant" %in% trt)) {
    # based on observations of Transplant in the dataset
    # we can remove it at the beginning, to make things easier for LOT counts
    trt <- trt[trt != "Transplant"]
  }  
  
  og_num <- length(trt) # number of lots in original (should be based on trt length)
  lines <- 1 # the first line is always the first lot
  trt_prev <- strsplit(trt[1], ",")[[1]]  # initialize first treatment
  
  for (i in 1:og_num) {
    trt2 <- strsplit(trt[i], ",")[[1]] # split up trt by "," (trt is a list)
    
    if (i != 1){ # make sure we don't go over the limit 

      # make the comparison using i - 1 
      trt1 <- strsplit(trt[i-1], ",")[[1]] # split up trt by ","
      
      # here we'll codify the rules
      
      # rule 1: discontinuing a previous line means new lot    
      if (any(trt2 %in% trt1)) { # observed a repeated drug in a trt (same lot)
        
        # rule 2: adding/substituting 1+ drugs means new lot
        if (any(!(trt2 %in% trt1))) { # observed something in trt2 not in trt1 (new lot)
          
          # however, if we add back something present before (look at trt_prev)
          # then that would not be a new lot
          if (!all(trt2 %in% trt_prev)) { # returns false if trt2 is a subset of trt_prev
            
            lines <- lines + 1 # save as an extra line
            trt_prev <- strsplit(trt[i], ",")[[1]]
            
          }
          
        }
        
      } else { # observed no repeated drug in a trt (new lot)
        
        lines <- lines + 1 # save as an extra line
        trt_prev <- strsplit(trt[i], ",")[[1]]
        
      }
      
    }
    
  }
  
  return(lines) # return the final number of LOT
  
}
```

```{r test_case1}
# to view a specific patient data 
df_patient[1, ]; df_patient[1, ]$trt

# run our algorithm on patient/row 1
lot_algorithm(df_patient[1, ]) # we obtain number of LOT = 13, which is correct 

# further examples: 
#lot_algorithm(df_patient[2, ])
#lot_algorithm(df_patient[3, ])
#lot_algorithm(df_patient[3, ])
#lot_algorithm(df_patient[110, ])
```

This code works successfully with the cases above. 
We will now write up some code to compare the LOT algorithm with the Reviewer's LOT for the remaining cases. 

\newpage

# Algorithm Performance

We run the algorithm on all cases and save the LOT in a column (lot_alg) of df_patient. 

```{r all_cases}

lot_alg <- c() # initialize a vector to save values from the algorithm

for (i in 1:nrow(df_patient)){
  lot_alg <- c(lot_alg, lot_algorithm(df_patient[i, ]))  
}

# add the vector as a column to df_patient
df_patient$lot_alg <- lot_alg

```

In the following code, we will compare the ability of the LOT algorithm to classify LOT compared to the Reviewer's LOT. In this dataset, Reviewer 1 and Reviewer 2 have the same LOT values, although I left both variables rev1_lot and rev2_lot in the data frame. However, I will make the comparison of the LOT algorithm with only one of the Reviewer's LOT. 

## Algorithm LOT Compared to Reviewer 1 LOT

```{r compare_case_alg}
# comparison vectors that return TRUE/FALSE for each row
alg_greater_than <- df_patient$lot_alg > df_patient$rev1_lot
alg_less_than <- df_patient$lot_alg < df_patient$rev1_lot
alg_equal_to <- df_patient$lot_alg == df_patient$rev1_lot

# View results (average of the TRUE/FALSE)
results <- round(c(sum(alg_greater_than) / nrow(df_patient), 
             sum(alg_less_than) / nrow(df_patient), 
             sum(alg_equal_to) / nrow(df_patient)), 3)

print(paste(results))  

```

Currently, we have 73.6% successful classification using our LOT algorithm, when compared to the Reviewer's LOT classification. 19.1% of the LOT classification from the algorithm are greater than the Reviewer's LOT classification. 7.3% of the LOT classification from the algorithm are smaller than the Reviewer's LOT classification.

## Dataset LOT Compared to Reviewer 1 LOT
```{r compare_case_data}
# comparison vectors that return TRUE/FALSE for each row
dat_greater_than <- df_patient$count_row > df_patient$rev1_lot
dat_less_than <- df_patient$count_row < df_patient$rev1_lot
dat_equal_to <- df_patient$count_row == df_patient$rev1_lot

# View results (average of the TRUE/FALSE)
results <- round(c(sum(dat_greater_than) / nrow(df_patient), 
             sum(dat_less_than) / nrow(df_patient), 
             sum(dat_equal_to) / nrow(df_patient)), 3)

print(paste(results))  
```

The original dataset's algorithm also has 73.6% successful classification when compared to the Reviewer's LOT classification. However they have a higher rate of over-counting LOTs (23.6%) than under-counting LOTs (2.7%). 

## Misclassified Cases

Following are the misclassified cases of the dataset, my LOT algorithm, and both the dataset and my algorithm.  

```{r wrong_cases}
compare_df <- data.frame(Row = integer(), Reviewer_LOT = integer(), 
                         Algorithm_LOT = integer(), Dataset_LOT = integer(), 
                         Error = character())

for (i in 1:nrow(df_patient)){
  if ((alg_equal_to[i] == FALSE) & (dat_equal_to[i] == FALSE) ) {
    compare_df <- compare_df %>% 
      add_row(Row = i, Reviewer_LOT = df_patient[i,]$rev1_lot, 
              Algorithm_LOT = df_patient[i,]$lot_alg, 
              Dataset_LOT = df_patient[i,]$count_row, Error = "Both")
    
  } else if ( (alg_equal_to[i] == TRUE) & (dat_equal_to[i] == FALSE) ) {
    compare_df <- compare_df %>% 
      add_row(Row = i, Reviewer_LOT = df_patient[i,]$rev1_lot, 
              Algorithm_LOT = df_patient[i,]$lot_alg, 
              Dataset_LOT = df_patient[i,]$count_row, Error = "Dataset") 
    
  } else if ( (alg_equal_to[i] == FALSE) & (dat_equal_to[i] == TRUE) ) {
    compare_df <- compare_df %>% 
      add_row(Row = i, Reviewer_LOT = df_patient[i,]$rev1_lot, 
              Algorithm_LOT = df_patient[i,]$lot_alg, 
              Dataset_LOT = df_patient[i,]$count_row, Error = "Algorithm")
    
  }
}

compare_df
```

I looked at a few of my misclassified cases to see where the error was. I think for most of these cases, manual review is needed to determine the true LOT.

- Patients 7 & 17 were misclassified due to "Line Zero". I didn't write this into the algorithm because that seems specific to this dataset. The reviewers classified it as 0 LOT because they were also unsure.

- Patient 8 was misclassified due to singular "Lenalidomide" not counting as a new LOT. This isn't always the case (e.g. Patient 1 has a singular Lenalidomide, but classifies a new LOT). 

```{r tester_Len}
len <- data.frame()

# observe what the singular Lenalidomide rows look like
for (i in 1:nrow(df_patient)) {
  patient <- df_patient[i, ]
  trt <- patient$trt[[1]] # extract vector of trt in a row
  
  if ("Lenalidomide" %in% trt) { # Lenalidomide administered alone
     len <- rbind(len, patient)
  }
  
}

head(len)
```

- Patient 21 was misclassified due to singular "Prednisone" not counting as a new LOT. This is the only case where it occurs in the data. 

```{r tester_pred}
pred <- data.frame()

# observe what the singular Prednisone rows look like
for (i in 1:nrow(df_patient)) {
  patient <- df_patient[i, ]
  trt <- patient$trt[[1]] # extract vector of trt in a row
  
  if ("Prednisone" %in% trt) { # Prednisone administered alone
     pred <- rbind(pred, patient)
  }
  
}

pred
```

- Patient 27 was misclassified when "Carfilzomib" was added back into the treatment plan. It was previously there, but after a gap. This one I was not able to account for in my algorithm, but it may be possible to add it.  

```{r test_case27}
df_patient[27, ]; df_patient[27, ]$trt
```

- Patient 28 was misclassified when "Daratumumab/Hyaluronidase-Fihj" was added to treatment containing other drugs, but not counted as a new LOT. 

```{r test_case28}
df_patient[28, ]; df_patient[28, ]$trt
```

# Next Steps

To further improve the algorithm, I would consider adding some more code that can highlight peculiar cases in the treatment regimens that should be manually reviewed (e.g. "Line Zero"). One idea may be to add a vector containing all drug/treatment names to help filter out unfamiliar terminology. 

Also, if my algorithms' accuracy could be further improved first, then maybe using a comparison to the  original dataset's LOT classification would be a good indicator for cases to be manually reviewed (currently their performance is similar so adding this indicator would not be very beneficial). 